State of the Practice for Lattice Boltzmann Method Software

### 2022-02-08 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Completed the edits of Section 5 (Comparison Between LBM and Research Software for Artifacts)

All of the top four AHP ranked packages, ESPREsSo, Ludwig, Palabos, and OpenLB,
have each of the commonly found artifacts, except Palabos does not have any
theory notes.  The top four AHP ranked packages have most of the less common
artifacts. At the time of data collection, only one of the four packages
(Palabos) did not have a user manual or guide, but there was a broken link on
the package website indicating that such an artifact might exist. This broken
link was later fixed, but this is not reflected in our data because it was not
present at the time of initial data collection (mid 2020). Despite the broken
link, Palabos does have a detailed and informative website. Another one (Ludwig)
of the top four packages does not appear to have publicly visible design
documentation. A third package (OpenLB) from the list does not appear to use a
version control system. It is possible that such a system is used by the
package, since its website notes package version numbers, but the artifacts do
not explicitly state the use of such a system.

@@

Although the LBM community participates in most of the practices we found listed
in general research software guidelines, some recommended practices were not observed or rarely observed. In addition to the items in the last column of Table~\ref{artifactspresent}, we can add the following community recommended items that we rarely, if ever, observed:

\begin{itemize}
\item A roadmap showing what is planned for the future \citep{Yo2021,
ThielEtAl2020, vanGompelEtAl2016}.  We did see this information mentioned in
some repos, but we did not observe a specific artifact devoted to this purpose.
\item A code of conduct to explicitly say how developers should treat one
another \citep{Yo2021, ThielEtAl2020}.  We did not observe a policy of this
sort, possibly because the number of developers for each LBM project is small.
\item Programming style guidelines so that programming labels and formatting are
consistent \citep{TobiasEtAl2018, ThielEtAl2020, OrvizEtAl2017, Zadka2018,
vanGompelEtAl2016, WilsonEtAl2014}.  We saw this information as part of
developer guides, but only rarely.
\item Checklists can be used in projects to ensure that best practices are
followed by all developers.  Some examples include checklists merging branches
into master \citep{Brown2015}, checklists for saving and sharing changes to the
project \citep{WilsonEtAl2016}, checklists for new and departing team members
\citep{HerouxAndBernholdt2018}, checklists for processes related to commits and
releases \citep{HerouxEtAl2008} and checklists for overall software quality
\citep{ThielEtAl2020, SSI2022}.  We did see evidence of checklists in the LBM
projects, but they were used only rarely.
\item Although not observed for LBM projects, having instructions to uninstall
the software is part of best practices \citep{vanGompelEtAl2016}.
\end{itemize}

The above discussion shows that, taken together, LBM projects fall short of
recommended practices for research software.  However, LBM software is not alone
in this.  Info on req, design and testing.  \citet{Nguyen-HoanEtAl2010} showed
requirements are the least commonly produced type of documentation for
scientific software in general.  Carver points out do not do requirements.  Not
everyone recommends requirements - can find citations.  Feasible to do
requirements.  Theory documents are already common (17 of 24? packages).  Need
to add just a little more information.

\wss{Document theory (17 of the packages), but not requirements.  Opportunity to
 go a little further and have a more complete requirements specificaiton. Add
 traceability to the code.  For instance, what are the consequences of changing
 an assumption?  adding an abstract transformation from theorty to code - when
 and how.  Example have a PDE, sate use finite difference. Consequence of change
 - likely and unlikely changes - give example of 2D vs 3D.  Basic idea is to
 give information on some simple things to add that would make a theory document
 like a requirements document. }

\wss{More on requirements as a pain point is given in Weiss, and other papers
cited by Carver (2021).  Isn't generally requirements in the SE sense of
following a template or similar.}

\wss{Weiss mentions pain points related to design documentation.  Carver (2021) 
also mentions design documentation in the literature review.}

In Table~\ref{artifactspresent} a rigorous requirements specification is rare,
but parts of a full specification do occur, with theory notes being common and
portions of the functional specification (statements about what the software
does) being less common.  Discuss requirements stuff.

@@

@@

\wss{Also compare to recommended practices for research software via the
checklists and other articles.}

>>>>

The top four packages, ESPREsSo, Ludwig, Palabos, and OpenLB, have all of the
common artifacts, except Palabos does not have theory notes.  The top four
packages also have most of the less common artifacts. At the time of data
collection, only one of the four packages (Palabos) did not have a user manual
or guide, but there was a broken link on the package website indicating that
such an artifact might exist. This broken link was later fixed, but this is not
reflected in our data because it was not present at the time of data collection
(mid 2020). Despite the broken link, Palabos does have a detailed and
informative website. As far as we could observe, Ludwig does not appear to have
publicly visible design and OpenLB does not appear to use a version control
system. However, OpenLB's use of version numbers suggests that version control
may be used privately by the developers.

@@

Although the LBM community participates in most of the practices we found listed
in the general research software guidelines, some recommended practices were not
observed or rarely observed. In addition to the items in the last column of
Table~\ref{artifactspresent}, we can add the following community recommended
items that we rarely, if ever, observed:

\begin{itemize}
\item A roadmap showing what is planned for the future \citep{Yo2021,
ThielEtAl2020, vanGompelEtAl2016}.  We did see this information mentioned in
some repos, but we did not observe a specific artifact devoted to this purpose.
\item A code of conduct to explicitly say how developers should treat one
another \citep{Yo2021, ThielEtAl2020}.  We did not observe a policy of this
sort, possibly because the number of developers for each LBM project is small.
\item Programming style guidelines so that programming labels and formatting are
consistent \citep{TobiasEtAl2018, ThielEtAl2020, OrvizEtAl2017, Zadka2018,
vanGompelEtAl2016, WilsonEtAl2014}.  We saw this information as part of
some developer guides, but only rarely.
\item Checklists can be used in projects to ensure that best practices are
followed by all developers.  Some examples include checklists merging branches
into master \citep{Brown2015}, checklists for saving and sharing changes to the
project \citep{WilsonEtAl2016}, checklists for new and departing team members
\citep{HerouxAndBernholdt2018}, checklists for processes related to commits and
releases \citep{HerouxEtAl2008} and checklists for overall software quality
\citep{ThielEtAl2020, SSI2022}.  For LBM software, ESPResSo has a checklist for
managing releases, but otherwise they are used only rarely for LBM.
\item Uninstall instructions \citep{vanGompelEtAl2016} were not observed for any
of the LBM projects.
\end{itemize}

The above discussion shows that, taken together, LBM projects fall short of
recommended best practices for research software.  However, LBM software is not
alone in this.  Many, if not most, research projects fall short of best
practices.  Software requirements documentation provides a good example of the
common deviation between recommendations and practice.  Although requirements
documentation is recommended by some \citep{TobiasEtAl2018, HerouxEtAl2008,
SmithAndKoothoor2016}, in practice scientific software developers often do not
produce a proper requirements specification \citep{HeatonAndCarver2015}.
\citet{SandersAndKelly2008} interviewed 16 scientists from 10 disciplines and
found that none of the scientists created requirements specifications, unless
regulations in their field mandated such a document. \citet{Nguyen-HoanEtAl2010}
showed requirements are the least commonly produced type of documentation for
scientific software in general. When looking at the pain points for research
software developers, \citet{WieseEtAl2019} found that software requirements and
management is the software engineering discipline that most hurts scientific
developers, accounting for 23 \% of the technical problems reported by study
participants.  The lack of support for requirements is likely due to the
perception that up-front requirements are impossible for research software
\citep{CarverEtAl2007,SegalAndMorris2008}, but when the instance on ``up-front''
requirements is dropped, allowing the requirements to be written iteratively and
incrementally, requirements are feasible \citep{Smith2016}.

Requirements documentation is feasible, as shown by the prevalence of theory
documents in LBM software.  Table~\ref{artifactspresent} shows that theory
documentation is common.  We observed 17 of 24 packages had at least some theory
documentation. Templates for scientific requirements, like \citet{SmithEtAl2007,
SmithAndLai2005}, show theory is a significant part of the requirements
documentation. With the addition of some extra information, the theory documents
can be transformed into requirements specification.  The key extra information
includes explicitly stating user characteristics, explicitly stating how the
user interacts with the software in terms of input data requirements, and
listing likely and unlikely changes.  Explicit statements about likely changes
are invaluable in the design stage, since they provide developers guidance on
how general the software needs to be.  For those wishing to maximize the value
of a requirement specification, information could be added like prioritizing the
nonfunctional requirements (to show the relative importance between qualities
like portability, reliability and performance) and traceability information (to
show the consequences of any changes to the assumptions).

@@

\wss{Beyond software carpenty has data on the increasing use of version control.}

@@
